Canonical rings of $\bq$-divisors on minimal rational surfaces

### 2016-03-19 surfaces-write-up/MRL/MRL-surfaces-responses-to-referee.tex

%        File: changes-summary.tex
%     Created: Sat Oct 17 11:00 AM 2015 E
% Last Change: Sat Oct 17 11:00 AM 2015 E
%
%%%%%%%%%%%%%%%%%%%%%
%   AMS packages    %
%%%%%%%%%%%%%%%%%%%%%
\documentclass[10 pt]{amsart}

\usepackage{amsmath}
\usepackage{amsxtra}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage[all]{xy}
\usepackage{graphicx}
\usepackage{tikz-cd}
\usepackage{mathrsfs}
\usepackage{subfiles}
\usepackage{mathpazo}
\usepackage{euler}
\usepackage[colorinlistoftodos, textsize=tiny]{todonotes}

\addtolength{\oddsidemargin}{-.5 in}
\addtolength{\evensidemargin}{-.4 in}
\addtolength{\textwidth}{1 in}
\addtolength{\topmargin}{0 in}
\addtolength{\textheight}{0 in}

\RequirePackage{color}
\definecolor{myred}{rgb}{0.75,0,0}
\definecolor{mygreen}{rgb}{0,0.5,0}
\definecolor{myblue}{rgb}{0,0,0.65}

\usepackage{hyperref}
\hypersetup{colorlinks=true,citecolor=blue}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows,decorations.pathmorphing}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{counterexample}[theorem]{Counterexample}
\newtheorem{convention}[theorem]{Convention}
\newtheorem{question}[theorem]{Question}
\newtheorem{conjecture}[theorem]{Conjecture} 
\newtheorem{warning}[theorem]{Warning}
\newtheorem{fact}[theorem]{Fact}
\theoremstyle{remark}
\newtheorem{notation}[theorem]{Notation}
\numberwithin{equation}{section}

\newcommand\nc{\newcommand}
\nc\on{\operatorname}
\nc\renc{\renewcommand}
\newcommand\se{\section}
\newcommand\ssec{\subsection}
\newcommand\sssec{\subsubsection}
\newcommand\bn{{\mathbb N}}
\newcommand\bc{{\mathbb C}}
\newcommand\br{{\mathbb R}}
\newcommand\bq{{\mathbb Q}}
\newcommand\bp{{\mathbb P}}
\newcommand\CF{{\mathcal F}}
\newcommand\bz{{\mathbb Z}}
\newcommand\ba{{\mathbb A}}
\newcommand\fa{{\mathfrak a}}
\newcommand\fp{{\mathfrak p}}
\newcommand\fq{{\mathfrak q}}
\newcommand\fm{{\mathfrak m}}
\newcommand\so{{\mathscr O}}
\newcommand\sg{{\mathscr G}}

\newcommand\scm{{\mathscr M}}
\newcommand\scn{{\mathscr N}}
\newcommand\scf{{\mathscr F}}
\newcommand\scg{{\mathscr G}}
\newcommand\sco{{\mathscr O}}
\newcommand\sch{{\mathscr H}}
\newcommand\scl{{\mathscr L}}
\newcommand\sci{{\mathscr I}}

\newcommand \ra{\rightarrow}
\newcommand{\id}{\mathrm{id}}
\newcommand\im{\text{im }}
\newcommand\coker{\text{coker}}
\newcommand \spec{\text{Spec }}
\newcommand \proj{\text{Proj }}
\newcommand \rspec{\textit{Spec }}
\newcommand \rproj{\textit{Proj }}
\newcommand \mg{{\mathscr M_g}}

\DeclareMathOperator\ord{ord}

\newcommand \trdeg{\text{tr. deg }}
\newcommand \codim{\text{codim}}
\newcommand \rk{\text{rk }}
\newcommand \di{\text{div }}
\newcommand \depth{\text{depth }}
\DeclareMathOperator\pic{Pic}
\DeclareMathOperator\lcm{lcm}
\DeclareMathOperator\rank{rank}
\DeclareMathOperator\vol{Vol}
\DeclareMathOperator\supp{Supp}

\def\listtodoname{List of Todos}
\def\listoftodos{\@starttoc{tdo}\listtodoname}

\title{Summary of Changes in response to Referee's Report}
\author{Aaron Landesman, Peter Ruhm, and Robin Zhang}

\begin{document}

\maketitle
\section{Changes Made}
Here are the changes we have made in response to the referee's suggestions.
\todo{make sure the theorem number here reflect the new numberings}
\begin{enumerate}
	\item \todo{Ask David how the new version sounds} We explained
		in more detail how our potential application
		for Hilbert modular forms and the difficulty in producing
		an explicit example.
	\item \todo{do this} For example, why give in Section 1.1 the main results in a form
The presentation can be improved quite a lot.
that the reader can only understand if he/she first reads later sections. 
\todo{Reorganize the intro so as to introduce notations and conventions first,
since the referee wants this}
	\item In all our main theorems, we now say that $c_i/k_i$ is in reduced form. 
	\item In Thm 1.2,  k is a field. Throughout our paper, the results hold for arbitrary fields, but we may assume the field is algebraically closed because the generator and relation degrees are preserved by base change to the algebraic closure. We have now included and clarified this in the ``Notations and Conventions'' subsection at the end of the Introduction.
	\item In the previous version we submitted, Page 2, line -9, $deg(D)$ should have been $\frac{1}{2 \prod_{i=0}^n q_i}$. This is now fixed.
		\todo{Say where}
	\item \todo{Make Thm. 1.4 readable by including necessary notations
		before it}
	\item Section 2 starts with definitions that should have come earlier. (Was Section
1.1 added in a later stage ?) Page 5, line 7, reference to things that come later.
\item In Thm 2.5 the best lower approximations are not explained. These come later
(page 9 and 10).
\item 
	\todo{I remember seeing something wrong with the bibliography, but can't find it now. Do you know what the referee is talking about?} The bibliography should be updated and misprints corrected.

\end{enumerate}

\end{document}

### 2016-03-19 surfaces-write-up/surface-write-up.tex
adding response to MRL referee and fixing many issues

\title{Canonical rings of $\bq$-divisors on Minimal Rational Surfaces}

@@

\begin{document}
\begin{abstract}
 	We give bounds on the degree of generation and relations of
	canonical rings of arbitrary $\bq$-divisors on projective spaces of 
	all dimensions and Hirzebruch surfaces. We also give exact bounds 
	on the degree of generators and relations of the canonical ring of 
	any effective stacky divisor on projective space.
\end{abstract}

@@

We expect the work in this paper will help calculate a presentation of certain rings of
Hilbert 
modular forms and Siegel modular forms. Recall that Hilbert and Siegel 
modular forms are two generalizations of modular forms to higher dimensions,
as described in \cite{geer:siegel-modular} and \cite{bruinier:hilbert-modular}.
Voight and Zureick-Brown show that 
one can translate the analytic category of 1-dimensional complex orbifolds to
the algebraic category of stacky curves \cite[Proposition 6.1.5]{vzb:stacky}. 
We expect that one can generalize this result to yield an equivalence of
categories between an appropriate subcategory of $m$-dimensional orbifolds and
a subcategory of $m$-dimensional algebraic spaces.
In particular, this would allow us to translate rings of Hilbert modular forms
and Siegel modular forms whose coarse space is a projective space or a
Hirzebruch surface to canonical rings on the corresponding algebraic surfaces.
Thereby, our techniques would give a bound on generators and relations for such
rings of Hilbert and Siegel modular forms.

@@

In Section \ref{sec:proj} we prove the following two results bounding
the degree of generators and relations of
stacky canonical rings on $\bp^m$. The first applies
to effective $\bq$-divisors and the second applies to arbitrary
$\bq$-divisors.
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{c_i}{k_i} \in \bq_{> 0}$ and each $D_i \in \di \bp^m$ an integral divisor.

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i = 0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$
with $\alpha_i = \frac{c_i}{k_i}\in \bq$. Write $\ell_i := \lcm
_{j \ne i}(k_j)$ and $a_i := \deg D_i$. Let $\bp^m \cong \proj
\bk[x_0, \ldots, x_m]$ and let each $D_i = V(f_i)$ with $f_i \in \bk[x_0, \ldots, 
x_m]$. Now suppose that $\{f_0, \ldots, f_n\}$ contains $m + 1$ independent 
linear polynomials in $x_0, \ldots, x_m$ {\rm(}with corresponding 
$\alpha_j$ possibly zero{\rm)}.

@@

\begin{rem}\label{rem:proj-tight-bounds} Note that the bounds
	given in Theorem ~\ref{thm:proj-effective-intro}
are tight and are typically attained, as explained in
Remark ~\ref{rem:exact-effective-bounds}.
Additionally, the bounds given in Theorem 
~\ref{thm:proj-generators-relations} are asymptotically
tight to within a factor of two 
for the following class of divisors. Consider
a divisor 
$ D = \sum_{i=0}^n\frac{p_i}{2q_i}H_i \in \bq \otimes_\bz \di \bp^m$
such that $H_i$ are hyperplanes,  
$q_i$ are pairwise coprime integers, and $p_i$ are chosen
so that $\deg D = \frac{1}{2 \prod_{i=0}^n}q_i$. Further,
choose a linear subspace $\pi: \bp^1 \rightarrow \bp^m$  generically so that
$\pi^* D = \sum_{i=0}^n \frac{p_i}{2q_i}P_i$, 
where $P_i$ are distinct points in $\bp^1$. To choose such a map
$\pi$, we may need to assume that the base field is infinite.
Note that by the Remark immediately following Theorem 8 in 
\cite{dorney:canonical},
the given bounds on the generators and relations of
$R_{\pi^*D}$ are within
a factor of two of the degree of generation and relations of
$R_{\pi^*D}$. Finally, since restriction map
$R_D \rightarrow R_{\pi^*D}$ induced by the restriction maps
on cohomology $H^0(\bp^m, dD) \rightarrow H^0(\bp^1, \pi^*(dD))
\cong H^0(\bp^1, d\pi^* D)$ are surjective, we obtain that
the bounds for the generators and relations of $R_D$ given in
Theorem ~\ref{thm:proj-generators-relations} also
agree with the degree of generation and relations 
to within a factor of two.
 \end{rem}

@@

\begin{restatable}{thm}{hirzrestate}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$.
Write $\ell_{i,j} := \lcm_{h\ne i,j}(k_h)$.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz} and define $\Te, \Tp$, and $\Tm$ as in Definition ~\ref{defn:t-defn}. Let each $D_i = V(f_i)$ where $f_i \in \sco(a_i, b_i)$.
Now suppose that $\{f_1, \ldots, f_n\}$ contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}with corresponding $\alpha_j$ possibly zero{\rm)}.

@@

\ssec{Notations and Conventions}
We will use $m \in \bz$ to index the dimension of a given projective
space $\bp^m$ or the type of the Hirzebruch surface $\hirz_m$.
Let $D$ be a (Weil) $\bq$-divisor of the form
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \bq \otimes_\bz \di X.
\end{align*}\noindent
where $n \in \bz$ indexes the number of irreducible divisors in 
the above expansion of $D$, $\alpha_i \in \bq$, and $D_i \in \di X$ is an integral
codimension 1 closed subscheme of $X$. When it is convenient to do
so, we shall sometimes start the indexing at $0$, so that $i$ runs
from $0$ to $n$. For such a divisor $D$ with $\alpha_i =
\frac{c_i}{k_i}$ let $\ell_i = \lcm_{j\ne i}(k_j)$
and let $\ell_{i,j} = \lcm_{h\ne i,j}(k_h).$ In the case $X = \bp^m$, define the degree of $D$ by $\deg D :=
\sum_{i=1}^{n}\alpha_i \cdot \deg D_i$. The floor of a $\bq$-divisor $D$ is a divisor $\lfloor D
\rfloor \in \di X$ is defined to be $\lfloor D \rfloor := \sum_{i = 1}^{n}
\lfloor \alpha_i \rfloor D_i$. Let $
	R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)$
denote the canonical ring associated to the $\bq$-divisor $D$. If $R$
is a graded ring, we denote the $d$th graded component of $R$ by $R_d$. If $r$ is a rational number, we let
$\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. Throughout, we
shall work over a fixed base field $\bk$. If $D \in \bq \otimes_\bz
\di X$ is an arbitrary divisor, we denote $h^0(X, D) := \dim_\bk H^0(X,D)$.

@@

\nocite{*}
\bibliography{bibliography-stacky-surface}{}
\bibliographystyle{plain}

>>>>

\title{Canonical rings of $\bq$-divisors on minimal rational surfaces}

@@

\begin{document}
\begin{abstract}
 	We give bounds on the degree of generation and relations of
	canonical rings of arbitrary $\bq$-divisors on projective spaces of 
	all dimensions and Hirzebruch surfaces. We also give exact bounds 
	on the degree of generators and relations of the canonical ring of 
	any effective $\bq$-divisor on projective space.
\end{abstract}

@@

We expect the work in this paper will help calculate a presentation of certain rings of
Hilbert 
modular forms and Siegel modular forms. Recall that Hilbert and Siegel 
modular forms are two generalizations of modular forms to higher dimensions,
as described in \cite{geer:siegel-modular} and \cite{bruinier:hilbert-modular}.
Voight and Zureick-Brown show that 
one can translate the analytic category of 1-dimensional complex orbifolds to
the algebraic category of stacky curves \cite[Proposition 6.1.5]{vzb:stacky}. 
We expect that one can generalize this result to yield an equivalence of
categories between an appropriate subcategory of $m$-dimensional orbifolds and
a subcategory of $m$-dimensional algebraic spaces.
In particular, this would allow us to translate rings of Hilbert modular forms
and Siegel modular forms whose coarse space is a projective space or a
Hirzebruch surface to canonical rings on the corresponding algebraic surfaces.
Thereby, if our results could be extended to all rational surfaces,
they would give a bound on generators and relations for such
rings of Hilbert and Siegel modular forms parameterized by rational
Hilbert and Siegel modular surfaces.
This would be interesting because such modular surfaces tend to be
immensely complicated: ``Already for $g = 2$, the boundary of this fundamental domain is complicated; Gottschling found that it possesses $28$ boundary pieces'' \cite[p.\ 7]{geer:siegel-modular}.
Since these varieties are so complicated, it is difficult
to give explicit examples in which we can completely work out the canonical
ring of a modular surface.
Although we did not obtain a bound on the degree of generators and relations
for general rational surfaces, its enough work to get a result for the class of rational varieties we consider.

@@

In Section \ref{sec:proj} we prove the following two results bounding
the degree of generators and relations of
stacky canonical rings on $\bp^m$. The first applies
to effective $\bq$-divisors and the second applies to arbitrary
$\bq$-divisors.
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{c_i}{k_i} \in \bq_{> 0}$ in reduced form and each $D_i \in \di \bp^m$ an integral divisor.

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i = 0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$
with $\alpha_i = \frac{c_i}{k_i}\in \bq$ in reduced form. Write $\ell_i := \lcm
_{j \ne i}(k_j)$ and $a_i := \deg D_i$. Let $\bp^m \cong \proj
\bk[x_0, \ldots, x_m]$ and let each $D_i = V(f_i)$ with $f_i \in \bk[x_0, \ldots, 
x_m]$. Now suppose that $\{f_0, \ldots, f_n\}$ contains $m + 1$ independent 
linear polynomials in $x_0, \ldots, x_m$ {\rm(}with corresponding 
$\alpha_j$ possibly zero{\rm)}.

@@

\begin{rem}\label{rem:proj-tight-bounds} Note that the bounds
	given in Theorem ~\ref{thm:proj-effective-intro}
are tight and are typically attained, as explained in
Remark ~\ref{rem:exact-effective-bounds}.
Additionally, the bounds given in Theorem 
~\ref{thm:proj-generators-relations} are asymptotically
tight to within a factor of two 
for the following class of divisors. Consider
a divisor 
$ D = \sum_{i=0}^n\frac{p_i}{2q_i}H_i \in \bq \otimes_\bz \di \bp^m$
such that $H_i$ are hyperplanes,  
$q_i$ are pairwise coprime integers, and $p_i$ are chosen
so that $\deg D = \frac{1}{2 \prod_{i=0}^n q_i}$. Further,
choose a linear subspace $\pi: \bp^1 \rightarrow \bp^m$  generically so that
$\pi^* D = \sum_{i=0}^n \frac{p_i}{2q_i}P_i$, 
where $P_i$ are distinct points in $\bp^1$. To choose such a map
$\pi$, we may need to assume that the base field is infinite.
Note that by the Remark immediately following Theorem 8 in 
\cite{dorney:canonical},
the given bounds on the generators and relations of
$R_{\pi^*D}$ are within
a factor of two of the degree of generation and relations of
$R_{\pi^*D}$. Finally, since restriction map
$R_D \rightarrow R_{\pi^*D}$ induced by the restriction maps
on cohomology $H^0(\bp^m, dD) \rightarrow H^0(\bp^1, \pi^*(dD))
\cong H^0(\bp^1, d\pi^* D)$ are surjective, we obtain that
the bounds for the generators and relations of $R_D$ given in
Theorem ~\ref{thm:proj-generators-relations} also
agree with the degree of generation and relations 
to within a factor of two.
 \end{rem}

@@

\begin{restatable}{thm}{hirzrestate}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$ is written in reduced form.
Write $\ell_{i,j} := \lcm_{h\ne i,j}(k_h)$.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz} and define $\Te, \Tp$, and $\Tm$ as in Definition ~\ref{defn:t-defn}. Let each $D_i = V(f_i)$ where $f_i \in \sco(a_i, b_i)$.
Now suppose that $\{f_1, \ldots, f_n\}$ contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}with corresponding $\alpha_j$ possibly zero{\rm)}.

@@

\ssec{Notations and Conventions}
We will use $m \in \bz$ to index the dimension of a given projective
space $\bp^m$ or the type of the Hirzebruch surface $\hirz_m$.
Let $D$ be a (Weil) $\bq$-divisor of the form
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \bq \otimes_\bz \di X.
\end{align*}\noindent
where $n \in \bz$ indexes the number of irreducible divisors in 
the above expansion of $D$, $\alpha_i \in \bq$, and $D_i \in \di X$ is an integral
codimension 1 closed subscheme of $X$. When it is convenient to do
so, we shall sometimes start the indexing at $0$, so that $i$ runs
from $0$ to $n$. For such a divisor $D$ with $\alpha_i =
\frac{c_i}{k_i}$ let $\ell_i = \lcm_{j\ne i}(k_j)$
and let $\ell_{i,j} = \lcm_{h\ne i,j}(k_h).$ In the case $X = \bp^m$, define the degree of $D$ by $\deg D :=
\sum_{i=1}^{n}\alpha_i \cdot \deg D_i$. The floor of a $\bq$-divisor $D$ is a divisor $\lfloor D
\rfloor \in \di X$ is defined to be $\lfloor D \rfloor := \sum_{i = 1}^{n}
\lfloor \alpha_i \rfloor D_i$. Let $
	R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)$
denote the canonical ring associated to the $\bq$-divisor $D$. If $R$
is a graded ring, we denote the $d$th graded component of $R$ by $R_d$. If $r$ is a rational number, we let
$\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. 
Throughout, we work over a fixed algebraically closed field $\bk$.
Since generator and relation degrees are preserved under base change to the algebraic closure, assuming that $\bk$ is algebraically closed is no restriction
on generality and the results hold equally well over an arbitrary field.

If $D \in \bq \otimes_\bz
\di X$ is an arbitrary divisor, we denote $h^0(X, D) := \dim_\bk H^0(X,D)$.

@@

\nocite{*}
\bibliography{bibliography-stacky-surface}
\bibliographystyle{alpha}
